The role of standing variation in geographic convergent adaptation

### 2015-08-20 standing-variation-paper.tex
minor grammar fixes from amnat

Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
If it does not encounter other locally fixed, beneficial mutants,
it will cover a distance $L$ in time $L/v$. In doing so (in two-dimensions) it will have
started from a point in space and spread out to cover a circular geographic region of area $\pi L^2$.
The cylinder in space-time with this circle as a base and
height $L/v$ has volume of this cylinder is $\pi L^2 \times (L/v)$.
Therefore, the number of other successfully established mutations that
would have appeared in the circle it has covered up until this time is Poisson with mean
$\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
(and $2 \lambda_0 L + 2 \lambda L^2 /v$ in one dimension).
% $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
\begin{equation} \label{eqn:chi_def_1}
    \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
\end{equation}
then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
before it is expected that one other successful mutation would have arisen in the area covered so far.
The explicit formula for $\chi$ in two dimensions can be found by
rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome; here we omit
it. In one dimension things are a little prettier and the
characteristic length is $\chi = ( \left(\lambda_0^2 + 2 \lambda/v \right) - \lambda_0 )^{1/2}/( 2 \lambda / v )$.
Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
in two-dimensions we can rewrite eqn. \eqref{eqn:chi_def_1} as
\begin{equation} \label{eqn:defines_chi}
%    \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^2 + \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^3 
%        = \frac{ - \log(1-s_d)^3 }{ 4 s_b^2 \sigma^2 \pi \rho \mu } .
   \frac{\sqrt{2s_b} }{-\log(1-s_d) } \chi^2 + \frac{1}{\sigma} \chi^3 = \frac{1}{\sqrt{2s_b} \rho\mu\pi}
\end{equation}
From this we see that $\chi$ decreases with $\rho$ and $\mu$.
Furthermore, for large $\sigma$, the characteristic length approaches
the value we would obtain just from standing variation:
\begin{equation} \label{eqn:chi_standing}
\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma).
\end{equation}
On the other hand, if the mutant allele is highly deleterious before $t=0$,
then standing variation is unimportant the characteristic length is approaches the value from \citet{ralphcoop2010}:
\begin{equation} \label{eqn:chi_new}
\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d)).
\end{equation}
These two end points help build our intuition for the interaction of
parameters in shaping the geographic scale of convergent evolution.
By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
 and occur within the first $\chi/v$ generations.
 Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
 we expect to find roughly one mutational origin.

@@

In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb}, we show the range of characteristic lengths as a function 
of various parameters chosen to match the evolution of malaria
resistance at G6PD. 
These curves match our intuition that higher population densities result in
smaller characteristic lengths (as would higher mutation rates). 
Allowing standing variation increases the 
input of new alleles and so decreases the characteristic length below that predicted
by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
In turn, increasing the prior deleterious effect of the allele ($s_d$)
acts to increase the
characteristic length until it reaches that predicted by new mutation alone. 
Higher dispersal distances lead to larger characteristic lengths, 
since more rapid geographic spread block other mutations from establishing. 
A larger selective advantage ($s_b$) acts in two conflicting
ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
to escape drift and become established. 
The effect of helping locally establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
 (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=s_b^{1/2}$,
showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
as in that case the speed of spread does not matter only the
  initial density of established alleles after the environmental shift. 
The dependence of the characteristic length on $s_b$ when only new
mutations contribute is much weaker (of order $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
Overall, the range of characteristic lengths observed are reasonably
consistent with the average diameter of a G6PD variant in Eurasia of
800km, especially for the lower population density, as long as the
fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was high.

Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
the characteristic length, is not
particularly intuitive we can use it to ask when we should expect
multiple adaptative alleles within a large geographic region  where our selective pressure
is present (thanks to Sam Yeaman for proposing this
interpretation). Consider the case where this geographic area is a fairly regular
shape of area $G$ and diameter $G^{1/2}$. Denote the total effective population size
over this area by $N = \rho G$, and the standard deviation of dispersal as a
fraction of the diameter of this area by $\sigma_G  = \sigma/
G^{1/2}$. Measuring distance in units of the diameter of $G^{1/2}$,
we expect multiple mutations when $\chi<1$ a
condition which will be met when 
\begin{equation}
N > \frac{\sigma_G}{ \mu \pi \sqrt{2s_b} \left(1 -  \sigma_G \sqrt{2s_b} /\log (1-s_d) \right)},
\end{equation}
this is found by rearranging eqn. \eqref{eqn:defines_chi} having set
$\chi=1$. This nicely shows that $\sigma_G/\mu$ is a primary
determinant of the critical population density necessary for
convergent evolution, when $\sigma_G$ is small. 
As $\sigma_G s_b/s_d$ becomes large, 
e.g., because the allele is not too deleterious before the environmental shift, 
this reduces to the situation with only standing variation,
in which the dependence of  $\sigma_G$
drops out, leaving the critical population size as $1/(2 \mu \pi s_b/s_d)$.

>>>>

Suppose we fix our attention on a particular new mutation that happens to be the first to occur in some region.
If it does not encounter other locally fixed, beneficial mutants,
it will cover a distance $L$ in time $L/v$. In doing so (in two-dimensions) it will have
started from a point in space and spread out to cover a circular geographic region of area $\pi L^2$.
The cylinder in space-time with this circle as a base and
height $L/v$ has a volume of $\pi L^2 \times (L/v)$.
Therefore, the number of other successfully established mutations that
would have appeared in the circle it has covered up until this time is Poisson with mean
$\lambda_0 \pi L^2 + \lambda \pi L^3 /v$ in two dimensions
(and $2 \lambda_0 L + 2 \lambda L^2 /v$ in one dimension).
% $\lambda_0 \omega_d L^d + \lambda \omega_d L^{d+1} /v$ in $d$ dimensions
Therefore, if we define $\chi$ for a two-dimensional model to be the unique positive solution to
\begin{equation} \label{eqn:chi_def_1}
    \lambda_0 \pi \chi^2 + \lambda \pi \chi^3 /v = 1,
\end{equation}
then $\chi$ gives the distance spread unobstructed by the descendants of a new mutant
before it is expected that one other successful mutation would have arisen in the area covered so far.
The explicit formula for $\chi$ in two dimensions can be found by
rearranging eqn. \eqref{eqn:chi_def_1} but is cumbersome; here we omit
it. In one dimension things are a little prettier and the
characteristic length is $\chi = ( \left(\lambda_0^2 + 2 \lambda/v \right) - \lambda_0 )^{1/2}/( 2 \lambda / v )$.
Substituting expressions for $\lambda_0$, $\lambda$, and $v$ from above,
in two-dimensions we can rewrite eqn. \eqref{eqn:chi_def_1} as
\begin{equation} \label{eqn:defines_chi}
%    \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^2 + \left( \frac{ - \chi \log(1-s_d) }{ 2 s_b \sigma } \right)^3 
%        = \frac{ - \log(1-s_d)^3 }{ 4 s_b^2 \sigma^2 \pi \rho \mu } .
   \frac{\sqrt{2s_b} }{-\log(1-s_d) } \chi^2 + \frac{1}{\sigma} \chi^3 = \frac{1}{\sqrt{2s_b} \rho\mu\pi}
\end{equation}
From this we see that $\chi$ decreases with $\rho$ and $\mu$.
Furthermore, for large $\sigma$, the characteristic length approaches
the value we would obtain just from standing variation:
\begin{equation} \label{eqn:chi_standing}
\chi = \sqrt{ -\log(1-s_d) / 2 s_b \rho \mu \pi } + O(1/\sigma).
\end{equation}
On the other hand, if the mutant allele is highly deleterious before $t=0$,
then standing variation is unimportant and the characteristic length is approaches the value from \citet{ralphcoop2010}:
\begin{equation} \label{eqn:chi_new}
\chi = ( \sigma / \rho \mu \pi \sqrt{2 s_b} )^{1/3} + O(1/\log(1-s_d)).
\end{equation}
These two end points help build our intuition for the interaction of
parameters in shaping the geographic scale of convergent evolution.
By the above calculation, we know that the relevant mutations occur about distance $\chi$ apart, 
 and occur within the first $\chi/v$ generations.
 Said another way, if we look in a circular region of space of radius $\chi$ over $\chi/v$ generations,
 we expect to find roughly one mutational origin.

@@

In Figures \ref{Fig-G6PD-charlength} and \ref{Fig-G6PD-charlength-sb}, we show the range of characteristic lengths as a function 
of various parameters chosen to match the evolution of malaria
resistance at G6PD. 
These curves match our intuition that higher population densities result in
smaller characteristic lengths (as would higher mutation rates). 
Allowing standing variation increases the 
input of new alleles and so decreases the characteristic length below that predicted
by only new mutations (i.e.\ equation \eqref{eqn:chi_new}, \citet{ralphcoop2010}).
In turn, increasing the prior deleterious effect of the allele ($s_d$)
acts to increase the
characteristic length until it reaches that predicted by new mutation alone. 
Higher dispersal distances lead to larger characteristic lengths, 
since more rapid geographic spread block other mutations from establishing. 
A larger selective advantage ($s_b$) acts in two conflicting
ways: aiding the rapid geographic spread of established alleles, and also helping more independent copies
to escape drift and become established. 
The effect of helping locally establish alleles wins out, since increasing the selective benefit $s_b$ decreases the characteristic length. 
 (This can be shown in general by differentiating \eqref{eqn:defines_chi} with respect to $z=s_b^{1/2}$,
showing that $\partial_{z} \chi = - ( C_1 z \chi + C_2 \chi^2 )/(C_3 z^2 + C_4 z \chi ) \le 0$ for appropriate nonnegative constants $C_{1-4}$.) 
This effect is strongest when only standing variation contributes ($s_b^{-1/2}$, equation \eqref{eqn:chi_standing}), 
as in that case only the initial density of established alleles after the environmental shift matters,
not the speed of spread.
The dependence of the characteristic length on $s_b$ when only new
mutations contribute is much weaker (of order $s_b^{-1/6}$, equation \eqref{eqn:chi_new})
Overall, the range of characteristic lengths observed are reasonably
consistent with the average diameter of a G6PD variant in Eurasia of
800km, especially for the lower population density, as long as the
fitness cost of G6PD-deficiency alleles before malaria ($s_d$) was high.

Finally, while the form of eqn. \eqref{eqn:defines_chi}, specifying
the characteristic length, is not
particularly intuitive we can use it to ask when we should expect
multiple adaptative alleles within a large geographic region  where our selective pressure
is present (thanks to Sam Yeaman for proposing this
interpretation). Consider the case where this geographic area is a fairly regular
shape of area $G$ and diameter $G^{1/2}$. Denote the total effective population size
over this area by $N = \rho G$, and the standard deviation of dispersal as a
fraction of the diameter of this area by $\sigma_G  = \sigma/
G^{1/2}$. Measuring distance in units of the diameter of $G^{1/2}$,
we expect multiple mutations when $\chi<1$, 
a condition which will be met when 
\begin{equation}
N > \frac{\sigma_G}{ \mu \pi \sqrt{2s_b} \left(1 -  \sigma_G \sqrt{2s_b} /\log (1-s_d) \right)},
\end{equation}
this is found by rearranging eqn. \eqref{eqn:defines_chi} having set
$\chi=1$. This nicely shows that $\sigma_G/\mu$ is a primary
determinant of the critical population density necessary for
convergent evolution, when $\sigma_G$ is small. 
As $\sigma_G s_b/s_d$ becomes large, 
e.g., because the allele is not too deleterious before the environmental shift, 
this reduces to the situation with only standing variation,
in which the dependence of  $\sigma_G$
drops out, leaving the critical population size as $1/(2 \mu \pi s_b/s_d)$.
